Computing MEM seeding probabilities

### 2019-02-08 MEM.tex
First non-draft version.

\begin{abstract}
The abstract will come later.
A popular mapping heuristic is called \emph{seeding}, and a recent variant
is the \emph{MEM seeding} approach. Here we tackle the question of how to
evaluate the probability that MEM seeding hits are on-target. This in turn
expresses the confidence one should have in the location of the mapped DNA
fragment and can be used to calibrate the MEM seeding heuristic, or to
filter the hits at the desired level of confidence.
\end{abstract}

@@

\section{The computational strategy}
\label{sec:symbolic}

@@

\subsection{Monte Carlo sampling}

@@

The reads are recoded as sequences of symbols from the following alphabet
$\mathcal{A}_* = \{\square, |, \downarrow_{/1}^-, \downarrow_{/2}^-,
\ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots, \Downarrow\}$. The
symbols $\downarrow_{/j}^-$ signify that the nucleotide is a mismatch
against the off-target sequence, the symbols $\downarrow_{/j}^+$ signify
that it is a mismatch against the target, and the symbol $\Downarrow$
signifies that it is a mismatch against both. As before, every other
nucleotide is replaced by the symbol $\square$, and the terminator $|$
is appended at the end of the read. We again define reads as sequences of
segments, except that now the terminators are the symbols
$\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.88]{sketch_dual.pdf}
\caption{\textbf{Title.}
Caption.}
\label{fig:dual}
\end{figure}

@@

\subsection{The probability of outcome $ii.$}

@@

\section{Conclusion}

This is awesome.

@@

We acknowledge the financial support of the Spanish Ministry of Economy
and Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan
Nacional BFU2012-37168), of the CERCA Programme~/~Generalitat de
Catalunya, and of the European Research Council (Synergy Grant 609989).

>>>>

\begin{abstract}
High throughput sequencing produces massive amounts of data that must be
processed efficiently. Most reads are identified by mapping them to a
reference, typically a genome. Modern mappers rely on heuristics to gain
speed at a reasonable cost of accuracy. MEM seeding is a popular heuristic
that shows good empirical performance but for which no theory is
available. It is thus difficult to calibrate or optimize mapping
strategies based on MEM seeding. Here we develop a theory to predict the
performance of MEM seeding. More specifically, we design a method to
compute the probability a read is mapped to the right location when using
MEM seeds. Our computational strategy is based on the analytic
combinatorics programme, but uses only part of it. By recoding the reads
in a specialized combinatorial alphabet, we can implicitly specify their
generating function and extract the probabilities of interest by direct
methods or by Monte Carlo sampling. This yields an efficient algorithm to
calibrate the MEM seeding heuristic for short read technologies.
\end{abstract}

@@

\section{Computational strategy}
\label{sec:symbolic}

@@

\subsection{Monte Carlo sampling}
\label{sec:montecarlo}

@@

The reads are recoded as sequences of symbols from the so-called ``+/-
alphabet'' $\mathcal{A}_{+/-} = \{\square, |, \downarrow_{/1}^-,
\downarrow_{/2}^-, \ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots,
\Downarrow\}$. The symbols $\downarrow_{/j}^-$ signify that the nucleotide
is a mismatch against the off-target sequence, the symbols
$\downarrow_{/j}^+$ signify that it is a mismatch against the target, and
the symbol $\Downarrow$ signifies that it is a mismatch against both. As
before, every other nucleotide is replaced by the symbol $\square$, and
the terminator $|$ is appended at the end of the read. We again define
reads as sequences of segments, except that now the terminators are the
symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.88]{sketch_dual.pdf}
\caption{\textbf{The +/- encoding.}
A example read is represented in the +/- alphabet. The symbols
$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
and the symbol $\Downarrow$ represent a mismatch against both. The index
$j$ is the number of matches since the last mismatch for the other
sequence. Note the presence of the terminator $|$ at the end of the read.
}
\label{fig:dual}
\end{figure}

@@

\subsection{The probability of false positives}

@@

\section{Discussion}

We have devised a method to compute the probability that the MEM seeding
heuristic misguides the mapping algorithm so that it cannot find the true
location of the read. The pillars of our strategy are borrowed from
analytic combinatorics~\cite{flajolet2009analytic,
sedgewick2013introduction}, even though we do not follow the standard
programme. Constructing generating functions usually serves the purpose of
finding their singularities in order to approximate the solution. In our
case, however, the weighted generating functions cannot be computed so
this strategy is not applicable.

To find the probabilities of interest in the absence of a fully specified
weighted generating function, we compute only the first terms using
iterative methods as in sections~\ref{sec:extract}, \ref{sec:cPE0} and
\ref{sec:cPE0E1}, or using Monte Carlo sampling as in
section~\ref{sec:montecarlo}. In this regard, the breakthrough is the
encoding of reads as MEM segments, which is a form of implicit Markov
imbedding~\cite{fu1994distribution}.

The method is general, but it is important to clearly state the
assumptions it depends on. First and most importantly, we have ignored
insertions and deletions. We assume that the sequencing errors are
substitutions only, which makes the method adapted to the Illumina
technology, but not to deletion-prone instruments such as the Oxford
Nanopore technology. We also assume that insertions and deletions never
occur among duplicated sequences. This is obviously incorrect, but our
initial tests with real data suggest that this is a minor impediment. On
real genomes, the estimations are only approximately correct and this is
but one of several simplifying assumptions. The theory is substantially
intrincate; insertions and deletions would make it intractable.
It is presently unclear how gracefully incorporate them in the theory.

The second assumption is that all the candidate locations with a MEM
seed of size $\geq \gamma$ are tested with the exact sequence alignment
algorithm, so that the best location is always discovered if it is in the
candidate set. This is not completely unrealistic, but it is important to
realize that for plant and animal genomes, a shared MEM seed may have tens
of thousands of hits. Therefore, most mappers impose a limit on the number
of alignments so the best hit may not be found even if it is in the
candidate set. This is again a minor impediment, since we can correct the
probability that the location is a false positive when only some
candidates have been tested.

The third assumption is that all the duplicate sequences evolve
independently of each other and at the same rate. This is again incorrect,
because duplication events can happen continuously, creating complex
ancestry relationships. It is possible to infer the ancestry using tree
reconstruction techniques, but it would be challenging to incorporate this
information in the present theory. The symbols of the MEM alphabet
implicitly assume that the sequences are exchangeable and the complexity
of the calculations explodes if it is not the case.

The last assumption is that MEM seeds can match only the target or one of
its duplicates. This is not the case: MEM seeds overlapping a sequencing
error often match random locations of the genome. When this occurs, the
MEM seeds are typically small and the alignment score at the candidate
location is very bad because the sequences are unrelated. So in practice,
it is relatively easy to detect this case and reject the location, even if
it is the best in the candidate set.

The proposed methods are compliant with the demand for speed of the
mapping algorithms. One needs to compute the probabilities only once for a
given value of $N$ and $\mu$ (the error rate $p$ is known and constant).
For $N > 20$, the iterative method is usually prohibitive and we need to
use Monte Carlo sampling instead, which for reads of size $< 200$
nucleotides keeps the computations around 1 second on modern hardware. The
values of $N$ and $\mu$ can be binned in intervals of interest so that
there are only $\approx 100$ pairs for a total cost of a few minutes per
run. Considering that mappers can process at most $10,000$ reads per
second, the computer time of mapping a sequencing run of 250 million reads
is over 7 hours, two orders of magnitude larger than the time required to
estimate the probabilities of error.

In conclusion, we presented a practical solution to the problem of
estimating the probability of false positives when using MEM seeds. This
solution is adapted for mapping short reads sequenced with the Illumina
technology. Being able to calibrate the MEM seeding heuristic not only
allows the user to choose how to balance speed versus accuracy, but also
opens new applications. For instance, one can map reads from contaminated
samples in pools of closely related genomes (\textit{e.g.}, modern human
and Neanderthal) in order to assign the reads to the organism they belong
to. The probabilities of false positives give the right level of
confidence in the assignment.

More generally, the analytic combinatorics programme is a very powerful
tool to address problems in bioinformatics. Here we have seen how this
strategy can be useful even when the generating functions cannot be
computed. Using the same ideas, one could calibrate heuristics used in
other alignment methods, especially in expanding fields such as long-read
technologies.

@@

We thank Ruggero Cortini for suggesting Monte Carlo methods as a way to
tackle intractable cases. We acknowledge the financial support of the
Spanish Ministry of Economy and Competitiveness (‘Centro de Excelencia
Severo Ochoa 2013-2017’, Plan Nacional BFU2012-37168), of the CERCA
Programme~/~Generalitat de Catalunya, and of the European Research Council
(Synergy Grant 609989).
